Efficient pedigree recording for fast population genetics simulation

### 2018-01-12 forwards_paper.tex
rearrangement

However, it is thought that most genetic variation is selectively neutral (or nearly so).
By definition, neutral alleles carried by individuals in a population
do not affect the population process.
For this reason, if one records the entire genealogical history of a population over the course of a simulation,
simply laying down neutral mutations on top of that history afterwards
is equivalent to having generated them during the simulation:
it does not matter if we generate each generation's mutations during the simulation, or afterwards.
Precisely, we need to know the genealogical trees relating all sampled individuals
at each position along the genome.
Combined with ancestral genotypes and the origins of new mutations,
these trees completely specify the genomic sequence of any individual in the population at any time.
To obtain this information, we record from forward simulation the \emph{population pedigree} --
the complete history of parent-offspring relationships of an entire population
going back to a remote time -- and the genetic outcomes of each ancestral meiosis,
periodically discarding all information irrelevant to the genetic history
of the extant population.
The information in this embellished pedigree is stored as a \emph{succinct tree sequence}
(or, for brevity, ``tree sequence''),
which contains all the information necessary
to construct the genealogical tree that relates each individual to every other
at each position on the genome.

@@

We now explain what we actually did to achieve a $50\times$ speedup.
The ``pedigree recording'' simulations above recorded information about each new individual
in a collection of tables that together define a \emph{succinct tree sequence}
(or, simply ``tree sequence'').
A {tree sequence} is an encoding for a sequence of correlated trees,
such as those describing the history of a sexual population.
Tree sequences are efficient because branches that are shared by adjacent trees are stored once,
rather than repeatedly for each tree.
The topology of a tree sequence is defined via its \emph{nodes} and \emph{edges},
while information about variants is recorded as \emph{sites} and \emph{mutations};
we give an example in Figure~\ref{fig:example_tree_sequence}.
This formulation is derived from the ``coalescence records'' encoding of tree
sequences~\citep{kelleher2016efficient}, normalised to remove redundancy
and generalised to include a more general class of tree topologies.

The idea of a tree sequence is closely related to the \emph{ancestral recombination graph},
or {ARG} \citep{griffiths1991two,griffiths1997ancestral},
which also describes the embellished pedigree.
The ARG has been the subject of substantial study
under the assumptions of coalescent
theory~\citep{wiuf1997number,wiuf1999ancestry,marjoram2006coalescent,wilton2015smc}.
However, the properties of the ARG as a computational structure have not
been studied and, despite several efforts to standardise a common
format~\citep{morin2006netgen,mcgill2013graphml}, % TODO check these refs against others in msprime paper
ARGs are rarely used in practise.
In contrast, the algorithmic properties of tree sequence
algorithms have been explored in detail~\citep{kelleher2016efficient},
contributing substantially to the efficiency of the \msprime{} coalescent simulator.

@@

The idea of storing genealogical information to speed up simulations is not new.
It was implemented completely in AnA-FiTS~\citep{aberer2013rapid},
but without the critical step of discarding irrelevant genealogical information.
A similar but much more limited method for
discarding this information also appears in \citet{padhukasahasram2008exploring}.

@@

Simulating very large populations and entire genomes will likely require parallelization.
The one-way nature of information flow here makes our scheme relatively easy to incorporate
into a parallel algorithm.
Recently, \citet{lawrie2017accelerating} used parallelization
to improve performance for the ``Poisson Random Field'' family of models~\citep{Sawyer1992-jw},
in which mutations are unlinked and there is no effect of genetic background on fitness.
However, thus far little attention has been paid to parallelization of more complex simulations.

\paragraph{Bayesian phylogenetics}
The methods described here for efficiently storing tree sequences may prove useful in other fields.
We have focused on the interpretation of tree sequences as the outcome of the process
of recombination, but in principle, we can efficiently encode any
sequence of trees which differ by subtree-prune-and-regraft
(SPR) operations. Since each SPR operation requires a constant amount of space to encode, the total
space required is $O(n + t)$ for $t$ trees with $n$ leaves~\citep{kelleher2016efficient}.
For instance, the large numbers of large, correlated trees produced by
MCMC samplers used in Bayesian phylogenetics \citep[e.g.,][]{drummond2012bayesian}
might be compactly stored as a tree sequence,
which would then allow highly efficient computation of properties of the posterior distribution.

>>>>

However, it is thought that most genetic variation is selectively neutral (or nearly so).
By definition, neutral alleles carried by individuals in a population
do not affect the population process.
For this reason, if one records the entire genealogical history of a population over the course of a simulation,
simply laying down neutral mutations on top of that history afterwards
is equivalent to having generated them during the simulation:
it does not matter if we generate each generation's mutations during the simulation, or afterwards.
To add mutations after the fact, we need to know the genealogical trees relating all sampled individuals
at each position along the genome.
Combined with ancestral genotypes and the origins of new mutations,
these trees completely specify the genomic sequence of any individual in the population at any time.
To obtain this information, we record from forward simulation the \emph{population pedigree} --
the complete history of parent-offspring relationships of an entire population
going back to a remote time -- and the genetic outcomes of each ancestral meiosis,
periodically discarding all information irrelevant to the genetic history
of the extant population.
The information in this embellished pedigree is stored as a \emph{succinct tree sequence}
(or, for brevity, ``tree sequence''),
which contains all the information necessary
to construct the genealogical tree that relates each individual to every other
at each position on the genome.

The idea of storing genealogical information to speed up simulations is not new.
It was implemented completely in AnA-FiTS~\citep{aberer2013rapid},
but without the critical step of discarding irrelevant genealogical information.
\citet{padhukasahasram2008exploring} obtained impressive speedups for a Wright--Fisher simulation
by keeping track of genealogies over the preceeding 8 generations
and only tracking neutral genotypes for those segments having descendants across this window.
Our approach is similar, but uses genealogies across the entire duration of the simulation.
The embellished pedigree is equivalent to the \emph{ancestral recombination graph},
or {ARG} \citep{griffiths1991two,griffiths1997ancestral},
which has been the subject of substantial study
% under the assumptions of coalescent theory
\citep{wiuf1997number,wiuf1999ancestry,marjoram2006coalescent,wilton2015smc}.
However, it is unclear if an ARG-based approach would share
the computational advantages of the data structures we use here \citet{kelleher2016efficient}.

% However, the properties of the ARG as a computational structure have not
% been studied and, despite several efforts to standardise a common
% format~\citep{morin2006netgen,mcgill2013graphml}, % TODO check these refs against others in msprime paper
% ARGs are rarely used in practise.
% In contrast, the algorithmic properties of tree sequence
% algorithms have been explored in detail~\citep{kelleher2016efficient},
% contributing substantially to the efficiency of the \msprime{} coalescent simulator.

@@

We now explain what we actually did to achieve this $50\times$ speedup.
The ``pedigree recording'' simulations above recorded information about each new individual
in a collection of tables that together define a \emph{succinct tree sequence}
(or, simply ``tree sequence'').
A {tree sequence} is an encoding for a sequence of correlated trees,
such as those describing the history of a sexual population.
Tree sequences are efficient because branches that are shared by adjacent trees are stored once,
rather than repeatedly for each tree.
The topology of a tree sequence is defined via its \emph{nodes} and \emph{edges},
while information about variants is recorded as \emph{sites} and \emph{mutations};
we give an example in Figure~\ref{fig:example_tree_sequence}.
This formulation is derived from the ``coalescence records'' encoding of tree
sequences~\citep{kelleher2016efficient}, normalised to remove redundancy
and generalised to include a more general class of tree topologies.

@@

@@

% Simulating very large populations and entire genomes will likely require parallelization.
% The one-way nature of information flow here makes our scheme relatively easy to incorporate
% into a parallel algorithm.
% Recently, \citet{lawrie2017accelerating} used parallelization
% to improve performance for the ``Poisson Random Field'' family of models~\citep{Sawyer1992-jw},
% in which mutations are unlinked and there is no effect of genetic background on fitness.
% However, thus far little attention has been paid to parallelization of more complex simulations.

\paragraph{Other applications}
The methods described here for efficiently storing tree sequences may prove useful in other fields.
We have focused on the interpretation of tree sequences as the outcome of the process
of recombination, but in principle, we can efficiently encode any
sequence of trees which differ by subtree-prune-and-regraft
operations. Since each such operation requires a constant amount of space to encode, the total
space required is $O(n + t)$ for $t$ trees with $n$ leaves~\citep{kelleher2016efficient}.
For instance, the large numbers of large, correlated trees produced by
MCMC samplers used in Bayesian phylogenetics \citep[e.g.,][]{drummond2012bayesian}
might be compactly stored as a tree sequence,
which would then allow highly efficient computation of properties of the posterior distribution.
